# Load TCM Data {#load-data}

`TCMDATA` provides two primary functions for retrieving Traditional Chinese Medicine data: `search_herb()` and `search_target()`.

Both functions return a `data.frame` containing three key columns:

- **herb**: Herb name
- **molecule**: Molecule/compound name  
- **target**: Target gene symbol

These columns represent the core components of TCM network pharmacology analysis.

---

## Search by herb name

Users can query TCM data by providing an herb name and specifying the name format:

```{r}
library(TCMDATA)

# Query using Chinese name
herbs <- c("灵芝")
lz <- search_herb(herb = herbs, type = "Herb_cn_name")
head(lz)
```

The `herb` parameter accepts three name formats:

| Type | Description | Example |
|------|-------------|---------|
| `Herb_cn_name` | Chinese characters (recommended) | `"灵芝"` |
| `Herb_pinyin_name` | Pinyin romanization (recommended) | `"lingzhi"` |
| `Herb_en_name` | English translation | `"Ganoderma"` or `"GANODERMA"` |

> 💡 **Tip**: Chinese and Pinyin formats are preferred for best search accuracy.

### Using Pinyin format

For Pinyin queries, concatenate all syllables without spaces or hyphens:

```{r}
# Correct: continuous pinyin
lz_py <- search_herb(herb = "lingzhi", type = "Herb_pinyin_name")
head(lz_py)
```

**Common mistakes to avoid:**
```{r, eval=FALSE}
# ✗ contains space
search_herb(herb = "ling zhi", type = "Herb_pinyin_name")

# ✗ contains hyphen
search_herb(herb = "ling-zhi", type = "Herb_pinyin_name")
```

---

## Search by target gene

All targets in `TCMDATA` are stored as gene symbols (e.g., HGNC nomenclature). You can retrieve associated herbs and molecules by querying target genes:

```{r}
# Query multiple target genes
genes <- c("TP53", "EGFR", "BRCA1")
results <- search_target(genes)
head(results)
```

The output shows which herbs and molecules interact with the specified targets, enabling reverse pharmacology analysis.

---

## Session information

```{r}
sessionInfo()
```
