# Molecule annotation {#molecule-detection}

After identifying candidate herbs and their associated molecules through the approaches described in `chapter 2`, it is often necessary to annotate these compounds with chemical descriptors and identifiers from public databases. `TCMDATA` provides a suite of PubChem<sup>[1]</sup> utilities to facilitate compound annotation and similarity searching.

---

## Resolving compound identifiers

The PubChem Compound Database assigns each chemical structure a unique identifier (CID). `TCMDATA` offers two functions to resolve molecule names or structural representations to CIDs.

### Using `resolve_cid()`

`resolve_cid()` supports multiple identifier types and uses the PubChem PUG-REST API directly. It accepts CIDs, SMILES strings, InChI, InChIKey, or compound names.

```{r, eval=FALSE, collapse=FALSE}
library(TCMDATA)

# Get CIDs for Lingzhi (Ganoderma) molecules
herbs <- c("灵芝")
lz <- search_herb(herb = herbs, type = "Herb_cn_name")

# Sample a few molecules
lz_mol <- head(unique(lz$molecule), 5)

# Resolve molecule names to CIDs
lz_mol_cid <- resolve_cid(lz_mol, from = "name")
print(lz_mol_cid)
```

```{r, eval=FALSE, collapse=FALSE}
# [1] "3,4-Dihydroxybenzoic acid" "4-Hydroxybenzoic Acid"     "Adenine"                  
# [4] "Uracil"                    "Thymidine" 
```

> 💡 **Note**: Check your network connection if you encounter errors during CID resolution.

The `from` argument specifies the input identifier type. Supported values include `"cid"`, `"smiles"`, `"inchi"`, `"inchikey"`, and `"name"`. For SMILES and InChI inputs, the function automatically handles URL encoding and uses POST requests to avoid length limitations.

### Using `getcid()`

`getcid()` wraps the `webchem::get_cid()` function and adds built-in rate limiting (≤5 requests per second) to comply with PubChem's usage policy. It is particularly useful when querying large batches of compound names.

```{r, eval=FALSE, collapse=FALSE}
# Alternative method using webchem backend
cid_results <- getcid(lz_mol, from = "name", match = "first")
```

---

## Retrieving chemical properties

Once CIDs are obtained, `getprops()` retrieves a range of chemical descriptors from PubChem, including molecular formula, molecular weight, IUPAC name, canonical SMILES, InChIKey, and LogP values.

```{r, eval=FALSE, collapse=FALSE}
# Get properties for resolved CIDs
props <- getprops(lz_mol_cid, properties = c("MolecularFormula", 
                                             "MolecularWeight", 
                                             "CanonicalSMILES", 
                                             "InChIKey"))

head(props)
```

```{r, eval=FALSE, collapse=FALSE}
# A tibble: 5 × 6
#  cid   CID   MolecularFormula MolecularWeight ConnectivitySMILES               InChIKey        
#  <chr> <chr> <chr>            <chr>           <chr>                            <chr>           
# 1 72    72    C7H6O4           154.12          C1=CC(=C(C=C1C(=O)O)O)O          YQUVCSBJEUQKSH-…
# 2 135   135   C7H6O3           138.12          C1=CC(=CC=C1C(=O)O)O             FJKROLUGYXJWQN-…
# 3 190   190   C5H5N5           135.13          C1=NC2=NC=NC(=C2N1)N             GFFGJBXGBJISGV-…
# 4 1174  1174  C4H4N2O2         112.09          C1=CNC(=O)NC1=O                  ISAKRJDGNUQOIC-…
# 5 5789  5789  C10H14N2O5       242.23          CC1=CN(C(=O)NC1=O)C2CC(C(O2)CO)O IQFYYKKMVGJFEH-…
```

The function internally uses rate limiting and retry logic to handle transient API errors. By default, it retrieves six common properties: `MolecularFormula`, `MolecularWeight`, `IUPACName`, `CanonicalSMILES`, `InChIKey`, and `XLogP`. Additional properties supported by PubChem can be specified via the `properties` argument.

---

## Similarity searching

Structural similarity searches are useful for identifying compounds with similar chemical scaffolds. `compound_similarity()` performs Tanimoto-based 2D similarity searches on PubChem using the Fast Similarity API.

```{r, eval=FALSE, collapse=FALSE}
# Search for compounds similar to Quercetin (CID 5280343)
sim_hits <- compound_similarity(
  query = "5280343",
  from = "cid",
  threshold = 90,
  topn = 10,
  compute_score = TRUE)

head(sim_hits)
```

```{r, eval=FALSE, collapse=FALSE}
# A tibble: 5 × 6
#  cid   CID   MolecularFormula MolecularWeight ConnectivitySMILES               InChIKey     
#  <chr> <chr> <chr>            <chr>           <chr>                            <chr>        
# 1 72    72    C7H6O4           154.12          C1=CC(=C(C=C1C(=O)O)O)O          YQUVCSBJEUQK…
# 2 135   135   C7H6O3           138.12          C1=CC(=CC=C1C(=O)O)O             FJKROLUGYXJW…
# 3 190   190   C5H5N5           135.13          C1=NC2=NC=NC(=C2N1)N             GFFGJBXGBJIS…
# 4 1174  1174  C4H4N2O2         112.09          C1=CNC(=O)NC1=O                  ISAKRJDGNUQO…
# 5 5789  5789  C10H14N2O5       242.23          CC1=CN(C(=O)NC1=O)C2CC(C(O2)CO)O IQFYYKKMVGJF…
```

The `threshold` parameter controls the minimum similarity percentage (0–100), and `topn` limits the number of returned hits. Setting `compute_score = TRUE` enables local Tanimoto score calculation using the `rcdk` package. This requires a working Java installation.

**Alternative query types:**

```{r, eval=FALSE, collapse=FALSE}
# Query by SMILES string
sim_by_smiles <- compound_similarity(
  query = "C1=CC(=C(C=C1O)O)C2=C(C(=O)C3=C(C=C(C=C3O2)O)O)O",
  from = "smiles",
  threshold = 85,
  topn = 5,
  compute_score = FALSE)
```

```{r, eval=FALSE, collapse=FALSE}
# Query by compound name
sim_by_name <- compound_similarity(
  query = "Quercetin",
  from = "name",
  threshold = 90,
  topn = 10,
  compute_score = FALSE)
```

The function returns a tibble with columns `query_cid`, `query`, `hit_cid`, `hit_smiles`, and `score` (if computed). This enables efficient filtering and prioritization of structurally related molecules for downstream pharmacological analysis.

---

## Download molecular structures and format conversion

To support downstream docking (such as AutoDock Vina) and structure-based analyses, `TCMDATA` also provides utilities for retrieving ligand and receptor structures and for converting molecular file formats. These functions are designed to streamline routine preprocessing steps and improve reproducibility in computational pharmacology workflows.

### Downloading ligand structures from PubChem

`download_ligand_structure()` retrieves SDF files for a given PubChem CID. By default (`type = "auto"`), the function prioritizes 3D conformers and automatically falls back to 2D records when 3D coordinates are unavailable.

Here, we download the structure for CID 2244 (Quercetin). If a curated 3D structure exists in PubChem, it will be downloaded; otherwise, the function will retrieve the 2D structure.
```{r, eval=FALSE, collapse=FALSE}
# Download ligand structure by CID (auto: 3D if available, otherwise 2D)
lig_sdf <- download_ligand_structure(
  cid = 2244,
  type = "auto",
  destdir = "structures/ligands",
  overwrite = FALSE)
```

This behavior is particularly useful in high-throughput settings, where compounds may differ in the availability of curated 3D structures.

### Downloading receptor structures from the RCSB PDB

`download_receptor_structure()` downloads macromolecular structures from the RCSB Protein Data Bank using a 4-character PDB identifier. The function supports both legacy PDB format (`format = "pdb"`) and PDBx/mmCIF format (`format = "cif"`).

Here, we download the structure for PDB ID 4hhb (human hemoglobin) in PDB format. 
```{r, eval=FALSE, collapse=FALSE}
# Download receptor structure by PDB ID
rec_pdb <- download_receptor_structure(
  pdb_id = "4hhb",
  format = "pdb",
  destdir = "structures/receptors",
  overwrite = FALSE)
```


### Converting structure file formats

`convert_structure()` performs general-purpose format conversion through Open Babel (`obabel`) and supports common structure formats including SDF, PDB, MOL2, XYZ, and SMILES. Optional arguments allow hydrogen addition and 3D coordinate generation during conversion.

Here, we demonstrate converting a ligand SDF file to PDB format while adding hydrogens. The `gen3d = FALSE` argument preserves existing coordinates if available.
```{r, eval=FALSE, collapse=FALSE}
# Convert ligand SDF to PDB and add hydrogens
lig_pdb <- convert_structure(
  input_file = lig_sdf,
  output_type = "pdb",
  add_hydrogens = TRUE,
  gen3d = FALSE,
  overwrite = TRUE)
```

### Recommended workflow

In a typical structure-preparation pipeline, users may: (i) obtain ligand CIDs via `resolve_cid()`/`getcid()`, (ii) download ligand SDF files from PubChem, (iii) download receptor structures from the RCSB PDB, and (iv) harmonize file formats for docking software using `convert_structure()`. This integrated workflow reduces manual preprocessing and promotes transparent, reproducible analyses.

---

## References

1. Sayers EW, Bolton EE, Fine AM, et al. (2026). "Database resources of the National Center for Biotechnology Information in 2026." *Nucleic Acids Research*, **54**(D1), D20–D27. doi: [10.1093/nar/gkaf1060](https://doi.org/10.1093/nar/gkaf1060).

---

## Session information

```{r, collapse=FALSE}
sessionInfo()
```
