# TCM network construction {#network-construction}

A core objective of network pharmacology is to reveal the multi-component, multi-target therapeutic mechanism of Traditional Chinese Medicine through network visualization. `TCMDATA` provides two complementary approaches for constructing and visualizing herb–molecule–target relationship networks:

1. **Sankey diagram** via `tcm_sankey()` — ideal for displaying flow-like, bipartite/tripartite relationships.
2. **Network graph** via `prepare_herb_graph()` + `ggtangle` — suited for topological analysis and node-level metric visualization.

---

## Preparing the data

All network visualizations start from the herb–molecule–target data frame returned by `search_herb()` or `search_target()`.

```{r net-data, collapse=FALSE}
library(TCMDATA)
library(dplyr)

# Query herbs and their molecular targets
herbs <- c("灵芝", "黄芪")
df <- search_herb(herb = herbs, type = "Herb_cn_name")

# Preview data dimensions
cat("Total rows:", nrow(df), "\n")
cat("Unique herbs:", n_distinct(df$herb), "\n")
cat("Unique molecules:", n_distinct(df$molecule), "\n")
cat("Unique targets:", n_distinct(df$target), "\n")
```

The returned data frame contains three core columns: `herb`, `molecule`, and `target`, representing the tripartite relationship.

---

## Sankey diagram

The Sankey diagram displays herb → molecule → target flows, with band width proportional to shared frequency. `tcm_sankey()` is powered by `ggalluvial` and supports rich color customization.

### Basic usage

```{r sankey-basic, fig.width=12, fig.height=9, fig.align='center', out.width='100%', collapse=FALSE}
library(ggplot2)

# Subset to top molecules (by frequency) for readability
top_mol <- df %>%
  count(molecule, sort = TRUE) %>%
  head(8) %>%
  pull(molecule)

df_sub <- df %>% filter(molecule %in% top_mol)

# Draw the Sankey diagram
p_sankey <- tcm_sankey(df_sub)
print(p_sankey)
```

Each vertical band represents a stratum (herb, molecule, or target), and the colored flows trace the associations between them. Wider bands indicate higher interaction frequency.

### Customizing colors and appearance

`tcm_sankey()` offers fine-grained control over color palettes, font style, flow transparency, and curvature:

```{r sankey-custom, fig.width=12, fig.height=9, fig.align='center', out.width='100%', collapse=FALSE}
p_sankey2 <- tcm_sankey(
  df_sub,
  herb_cols   = c("#E41A1C", "#377EB8"),
  mol_cols    = c("#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3",
                  "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3"),
  target_cols = c("#1B9E77", "#D95F02", "#7570B3", "#E7298A",
                  "#66A61E", "#E6AB02", "#A6761D", "#666666"),
  font_size   = 3.2,
  alpha       = 0.35,
  knot.pos    = 0.25
)
print(p_sankey2)
```

Key parameters:

| Parameter | Description |
|-----------|-------------|
| `herb_cols`, `mol_cols`, `target_cols` | Color vectors for each layer |
| `font_size` | Node label size |
| `alpha` | Flow transparency (0–1) |
| `knot.pos` | Curvature of flow bands (0–1) |
| `width` | Node/flow band width |

---

## Network graph with `ggtangle`

While the Sankey diagram excels at showing flows, a force-directed network layout better reveals topological structure, hub nodes, and community patterns. `prepare_herb_graph()` builds an `igraph` object with pre-computed node metrics, and [ggtangle](https://github.com/YuLab-SMU/ggtangle) renders it as a `ggplot`-based network.

### Building the graph

```{r build-graph, collapse=FALSE}
library(igraph)

# Build the network (using a moderate subset for clarity)
g <- prepare_herb_graph(df, n = 60, compute_metrics = TRUE)

# Inspect graph summary
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Node types:", paste(unique(V(g)$type), collapse = ", "), "\n")
```

`prepare_herb_graph()` automatically:

- Creates directed edges: Herb → Molecule → Target
- Computes per-node metrics: `degree`, `centrality` (eigenvector), `betweenness`, `closeness`, `pagerank`
- Assigns `type` labels (`"Herb"`, `"Molecule"`, `"Target"`) to each node

### Basic network visualization

`ggtangle` extends `ggplot2` to natively support `igraph` objects. You can use familiar aesthetics (`color`, `size`, `shape`) and layers (`geom_point`, `geom_text_repel`).

```{r net-basic, fig.width=10, fig.height=8, fig.align='center', out.width='95%', collapse=FALSE}
library(ggtangle)
library(ggrepel)

set.seed(42)
p_net <- ggplot(g, layout = "kk") +
  geom_edge(alpha = 0.3, color = "grey60") +
  geom_point(aes(color = type, size = centrality)) +
  scale_color_manual(values = c(
    "Herb"     = "#E41A1C",
    "Molecule" = "#377EB8",
    "Target"   = "#4DAF4A"
  )) +
  scale_size_continuous(range = c(2, 8)) +
  theme_void() +
  labs(title = "Herb–Molecule–Target Network",
       subtitle = "Layout: Kamada-Kawai | Node size: Eigenvector centrality")

print(p_net)
```

In this plot:

- **Red nodes** represent herbs, **blue** represent molecules, and **green** represent targets.
- **Larger nodes** have higher eigenvector centrality, indicating greater topological importance.
- The Kamada-Kawai (`"kk"`) layout positions strongly connected nodes closer together.

### Adding node labels

For networks with many nodes, `ggrepel::geom_text_repel()` prevents label overlap. Here we label only high-centrality nodes:

```{r net-labels, fig.width=11, fig.height=9, fig.align='center', out.width='95%', collapse=FALSE}
set.seed(42)

# Extract centrality threshold (top 30%)
cent_thresh <- quantile(V(g)$centrality, 0.7)

p_labeled <- ggplot(g, layout = "kk") +
  geom_edge(alpha = 0.25, color = "grey70") +
  geom_point(aes(color = type, size = centrality), alpha = 0.85) +
  geom_text_repel(
    aes(label = ifelse(centrality >= cent_thresh, name, "")),
    size = 3, max.overlaps = 25, segment.alpha = 0.4
  ) +
  scale_color_manual(values = c(
    "Herb"     = "#E41A1C",
    "Molecule" = "#377EB8",
    "Target"   = "#4DAF4A"
  )) +
  scale_size_continuous(range = c(1.5, 9)) +
  theme_void() +
  labs(title = "Labeled Network (Top 30% Centrality Nodes)")

print(p_labeled)
```

### Alternative layouts

`ggtangle` supports all `igraph` layout algorithms. Common choices include:

```{r net-layouts, fig.width=14, fig.height=10, fig.align='center', out.width='100%', collapse=FALSE}
library(aplot)

set.seed(42)
base_layers <- list(
  geom_edge(alpha = 0.25, color = "grey70"),
  geom_point(aes(color = type, size = centrality)),
  scale_color_manual(values = c(
    "Herb" = "#E41A1C", "Molecule" = "#377EB8", "Target" = "#4DAF4A")),
  scale_size_continuous(range = c(1.5, 7)),
  theme_void()
)

p_fr  <- ggplot(g, layout = "fr") + base_layers + ggtitle("Fruchterman-Reingold")
p_kk  <- ggplot(g, layout = "kk") + base_layers + ggtitle("Kamada-Kawai")
p_star <- ggplot(g, layout = "star") + base_layers + ggtitle("Star")
p_nicely <- ggplot(g, layout = "nicely") + base_layers + ggtitle("Nicely (auto)")

plot_list(p_fr, p_kk, p_star, p_nicely, ncol = 2)
```

| Layout | Best for |
|--------|----------|
| `"kk"` (Kamada-Kawai) | Medium-sized networks with clear clusters |
| `"fr"` (Fruchterman-Reingold) | Dense networks; force-directed separation |
| `"star"` | Highlighting a central hub node |
| `"nicely"` | Automatic layout selection by igraph |

### Enriching the network with external data

Using the `%<+%` operator from `ggfun`, you can bind additional data to the network for richer visual encodings:

```{r net-external, fig.width=10, fig.height=8, fig.align='center', out.width='95%', collapse=FALSE}
# Create a data frame with node-level metrics
node_df <- data.frame(
  label       = V(g)$name,
  type        = V(g)$type,
  degree      = V(g)$degree,
  betweenness = V(g)$betweenness,
  pagerank    = V(g)$pagerank
)

set.seed(42)
p_enriched <- ggplot(g, layout = "kk") +
  geom_edge(alpha = 0.2, color = "grey70") +
  geom_point(aes(color = type, size = degree), alpha = 0.85) +
  scale_color_manual(values = c(
    "Herb" = "#E41A1C", "Molecule" = "#377EB8", "Target" = "#4DAF4A"
  )) +
  scale_size_continuous(name = "Degree", range = c(2, 10)) +
  theme_void() +
  labs(title = "Network Colored by Node Type, Sized by Degree")

print(p_enriched)
```

---

## Comparing the two approaches

| Feature | Sankey (`tcm_sankey`) | Network (`ggtangle`) |
|---------|----------------------|----------------------|
| Best for | Displaying flows and proportions | Topology and hub detection |
| Layout | Fixed three-column | Flexible (force-directed, circular, etc.) |
| Node metrics | Not applicable | Degree, centrality, betweenness, PageRank |
| Interactivity | Static | Extensible with `ggiraph` |
| Scalability | ~50–200 connections | ~30–500 nodes |

Both visualizations are complementary: the Sankey diagram provides an intuitive overview of which herbs contribute which molecules to which targets, while the network graph reveals structural properties like hub molecules, bridge targets, and community modules.

---

## Session information

```{r, collapse=FALSE}
sessionInfo()
```