# TCM network construction {#network-construction}

A core objective of network pharmacology is to reveal the multi-component, multi-target therapeutic mechanism of Traditional Chinese Medicine through network visualization. `TCMDATA` provides two complementary approaches for constructing and visualizing herb–molecule–target relationship networks: **Sankey diagrams** via `ggalluvial`<sup>[1]</sup> for depicting the directional flow of interactions, and **force-directed network graphs** via `ggtangle`<sup>[2]</sup> for revealing topological structure and hub nodes.

---

## Preparing the data

All network visualizations start from the herb–molecule–target data frame returned by `search_herb()` or `search_target()`.

```{r net-data, collapse=FALSE}
library(TCMDATA)
library(dplyr)

# Query herbs and their molecular targets
herbs <- c("灵芝", "黄芪")
df <- search_herb(herb = herbs, type = "Herb_cn_name")

df |> head(10)
```

The returned data frame contains three core columns: `herb`, `molecule`, and `target`, representing the tripartite relationship.

---

## Sankey diagram visualization

The Sankey diagram displays herb → molecule → target flows, with band width proportional to shared frequency. `tcm_sankey()` is powered by `ggalluvial`<sup>[1]</sup> and supports rich color customization.

### Basic usage

```{r sankey-basic, fig.width=12, fig.height=18, fig.align='center', out.width='100%', collapse=FALSE}
library(ggplot2)

## prepare input data
top_mol <- df %>%
  count(molecule, sort = TRUE) %>%
  head(5) %>%
  pull(molecule)

df_sub <- df %>% filter(molecule %in% top_mol)

set.seed(2026)
sampled_targets <- df_sub %>%
  distinct(target) %>%
  sample_n(min(30, n())) %>%
  pull(target)

df_sankey <- df_sub %>% filter(target %in% sampled_targets)

p_sankey <- tcm_sankey(df_sankey)
print(p_sankey)
```

Each vertical band represents a stratum (herb, molecule, or target), and the colored flows trace the associations between them. Wider bands indicate higher interaction frequency.

### Customizing colors and appearance

`tcm_sankey()` offers fine-grained control over color palettes, font style, flow transparency, and curvature. Below is an example with a publication-ready color scheme:

```{r sankey-custom, fig.width=12, fig.height=18, fig.align='center', out.width='100%', collapse=FALSE}
p_sankey2 <- tcm_sankey(
  df_sankey,
  herb_cols   = c("#C44E52", "#4C72B0"),
  mol_cols    = c("#55A868", "#DD8452", "#8172B3", "#C44E52",
                  "#DA8BC3"),
  target_cols = c("#1A7A6D", "#BF6D2A", "#6A5ACD", "#C04779",
                  "#5B8C2A", "#D4A017", "#8B6914", "#607D8B"),
  font_size   = 3.5,
  alpha       = 0.4,
  knot.pos    = 0.25
)
print(p_sankey2)
```

---

## Network graph with `ggtangle`

While the Sankey diagram excels at showing flows, a force-directed network layout can reveal topological structure, hub nodes, and community patterns. `prepare_herb_graph()` builds an `igraph` object with pre-computed node metrics, and `ggtangle`<sup>[2]</sup> renders it as a `ggplot`-based network.

### Building the graph

```{r build-graph, collapse=FALSE}
library(igraph)

# Build the network (using a moderate subset for clarity)
g <- prepare_herb_graph(df, n = 60, compute_metrics = TRUE)

# Inspect graph summary
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Node types:", paste(unique(V(g)$type), collapse = ", "), "\n")
```

`prepare_herb_graph()` automatically:

- Creates directed edges: Herb → Molecule → Target
- Computes per-node metrics: `degree`, `centrality` (eigenvector), `betweenness`, `closeness`, `pagerank`
- Assigns `type` labels (`"Herb"`, `"Molecule"`, `"Target"`) to each node

### Basic network visualization

`ggtangle`<sup>[2]</sup> extends `ggplot2` to natively support `igraph` objects. You can use familiar aesthetics (`color`, `size`, `shape`) and layers (`geom_point`, `geom_text_repel`).

```{r net-basic, fig.width=10, fig.height=8, fig.align='center', out.width='95%', collapse=FALSE}
library(ggtangle)
library(ggrepel)

# Define a publication-quality color palette
node_colors <- c(
  "Herb"     = "#C44E52",
  "Molecule" = "#4C72B0",
  "Target"   = "#55A868"
)

# Define node shape mapping
node_shapes <- c(
  "Herb"     = 18,   # diamond
  "Molecule" = 16,   # filled circle
  "Target"   = 15    # filled square
)

set.seed(42)
p_net <- ggplot(g, layout = "kk") +
  geom_edge(alpha = 0.18, color = "grey60", linewidth = 0.3) +
  geom_point(aes(color = type, size = centrality, shape = type), alpha = 0.85) +
  scale_color_manual(values = node_colors, name = "Node Type") +
  scale_shape_manual(values = node_shapes, name = "Node Type") +
  scale_size_continuous(name = "Centrality", range = c(2, 9)) +
  guides(
    color = guide_legend(order = 1, override.aes = list(size = 4)),
    shape = guide_legend(order = 1),
    size  = guide_legend(order = 2)
  ) +
  theme_void() +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 10),
    legend.title     = element_text(size = 11, face = "bold"),
    plot.margin      = margin(5, 5, 5, 5)
  )

print(p_net)
```

In this plot:

- **Diamond** nodes (<span style="color:#C44E52">■</span>) represent herbs, **circles** (<span style="color:#4C72B0">●</span>) represent molecules, and **squares** (<span style="color:#55A868">■</span>) represent targets.
- **Larger nodes** have higher eigenvector centrality, indicating greater topological importance.
- The Kamada-Kawai (`"kk"`) layout positions strongly connected nodes closer together.

### Adding node labels

For networks with many nodes, `ggrepel::geom_text_repel()` prevents label overlap. Here we label only high-centrality nodes (top 50%):

```{r net-labels, fig.width=11, fig.height=9, fig.align='center', out.width='95%', collapse=FALSE}
set.seed(42)

# Extract centrality threshold (top 50%)
cent_thresh <- quantile(V(g)$centrality, 0.5)

p_labeled <- ggplot(g, layout = "kk") +
  geom_edge(alpha = 0.15, color = "grey55", linewidth = 0.3) +
  geom_point(aes(color = type, size = centrality, shape = type), alpha = 0.85) +
  geom_text_repel(
    aes(label = ifelse(centrality >= cent_thresh, name, "")),
    size = 2.8, fontface = "italic",
    max.overlaps = 30, segment.alpha = 0.3,
    segment.size = 0.3, segment.color = "grey50",
    box.padding = 0.4, point.padding = 0.3
  ) +
  scale_color_manual(values = node_colors, name = "Node Type") +
  scale_shape_manual(values = node_shapes, name = "Node Type") +
  scale_size_continuous(name = "Centrality", range = c(1.5, 10)) +
  guides(
    color = guide_legend(order = 1, override.aes = list(size = 4)),
    shape = guide_legend(order = 1),
    size  = guide_legend(order = 2)
  ) +
  theme_void() +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 10),
    legend.title     = element_text(size = 11, face = "bold"),
    plot.margin      = margin(5, 5, 5, 5)
  )

print(p_labeled)
```

### Alternative layouts

`ggtangle` supports all `igraph` layout algorithms. Common choices include:

```{r net-layouts, fig.width=14, fig.height=10, fig.align='center', out.width='100%', collapse=FALSE}
library(aplot)

set.seed(42)
base_layers <- list(
  geom_edge(alpha = 0.15, color = "grey55", linewidth = 0.3),
  geom_point(aes(color = type, size = centrality, shape = type), alpha = 0.85),
  scale_color_manual(values = node_colors, name = "Node Type"),
  scale_shape_manual(values = node_shapes, name = "Node Type"),
  scale_size_continuous(name = "Centrality", range = c(1.5, 7)),
  guides(
    color = guide_legend(order = 1, override.aes = list(size = 3.5)),
    shape = guide_legend(order = 1),
    size  = guide_legend(order = 2)
  ),
  theme_void(),
  theme(
    plot.title       = element_text(hjust = 0.5, face = "bold", size = 13),
    legend.position  = "bottom",
    legend.text      = element_text(size = 9),
    legend.title     = element_text(size = 10, face = "bold")
  )
)

p_fr     <- ggplot(g, layout = "fr")     + base_layers + ggtitle("Fruchterman–Reingold")
p_kk     <- ggplot(g, layout = "kk")     + base_layers + ggtitle("Kamada–Kawai")
p_star   <- ggplot(g, layout = "star")   + base_layers + ggtitle("Star")
p_nicely <- ggplot(g, layout = "nicely") + base_layers + ggtitle("Nicely (auto)")

plot_list(p_fr, p_kk, p_star, p_nicely, ncol = 2)
```


### Enriching the network with external data

Using the `%<+%` operator from `ggfun`, you can bind additional data to the network for richer visual encodings. Here, node size is mapped to **degree** (the number of direct interactions), and all node labels are displayed:

```{r net-external, fig.width=12, fig.height=10, fig.align='center', out.width='100%', collapse=FALSE}
# Create a data frame with node-level metrics
node_df <- data.frame(
  label       = V(g)$name,
  type        = V(g)$type,
  degree      = V(g)$degree,
  betweenness = V(g)$betweenness,
  pagerank    = V(g)$pagerank
)

set.seed(42)
p_enriched <- ggplot(g, layout = "kk") +
  geom_edge(alpha = 0.15, color = "grey55", linewidth = 0.3) +
  geom_point(aes(color = type, size = degree, shape = type), alpha = 0.85) +
  geom_text_repel(
    aes(label = name, color = type),
    size = 2.5, fontface = "italic",
    max.overlaps = 100, segment.alpha = 0.25,
    segment.size = 0.2, segment.color = "grey60",
    box.padding = 0.3, point.padding = 0.2,
    show.legend = FALSE
  ) +
  scale_color_manual(values = node_colors, name = "Node Type") +
  scale_shape_manual(values = node_shapes, name = "Node Type") +
  scale_size_continuous(name = "Degree", range = c(2, 11)) +
  guides(
    color = guide_legend(order = 1, override.aes = list(size = 4)),
    shape = guide_legend(order = 1),
    size  = guide_legend(order = 2)
  ) +
  theme_void() +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 10),
    legend.title     = element_text(size = 11, face = "bold"),
    plot.margin      = margin(10, 10, 10, 10)
  )

print(p_enriched)
```

---

## References

1. Brunson JC (2020). "ggalluvial: Layered Grammar for Alluvial Plots." *Journal of Open Source Software*, **5**(49), 2017. doi: [10.21105/joss.02017](https://doi.org/10.21105/joss.02017).

2. Yu G (2025). *ggtangle: Draw Network with Data*. R package version 0.1.1. doi: [10.32614/CRAN.package.ggtangle](https://doi.org/10.32614/CRAN.package.ggtangle).

---

## Session information

```{r, collapse=FALSE}
sessionInfo()
```