# Enrichment analysis {#enrichment}

`TCMDATA` provides functions for performing enrichment analysis on gene sets derived from herb–target interactions. It supports both Over-Representation Analysis (ORA) and Gene Set Enrichment Analysis (GSEA) methodologies via `clusterProfiler`, and offers a suite of visualization methods for interpreting enrichment results through `enrichplot` and other compatible packages.

## Over Representation Analysis (ORA)

ORA evaluates whether a pre-defined gene set (e.g., KEGG pathways, GO terms, Reactome pathways) is statistically over-represented among a list of input genes, such as the intersection of DEGs and disease targets. For a comprehensive treatment of enrichment methods, refer to the [*Biomedical Knowledge Mining* book](https://yulab-smu.top/biomedical-knowledge-mining-book/02-Enrichment.html).

### KEGG enrichment analysis

[KEGG](https://www.kegg.jp/) (Kyoto Encyclopedia of Genes and Genomes) is a comprehensive database that maps genes to metabolic pathways, signaling cascades, and other molecular interaction networks. KEGG pathway enrichment analysis identifies which biological pathways are significantly over-represented in a given gene set, thereby revealing the functional mechanisms underlying the observed gene expression changes.

In this example, we randomly sampled 200 targets of Lingzhi (*Ganoderma lucidum*) as input for KEGG pathway enrichment. Since `enrichKEGG()` requires **Entrez IDs** by default, we first converted gene symbols using `bitr()`:

```{r kegg-input, collapse=FALSE, message=FALSE, warning=FALSE}
library(TCMDATA)
set.seed(2026)

lz_targets <- search_herb("lingzhi", type = "Herb_pinyin_name")$target
query_genes <- sample(lz_targets, 200)

head(query_genes)
```

```{r kegg-convert, collapse=FALSE, message=FALSE, warning=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)

## convert Symbol to Entrez ID
gene_id <- bitr(query_genes,
                fromType = "SYMBOL",
                toType   = "ENTREZID",
                OrgDb    = org.Hs.eg.db)

head(gene_id)
```

Then, we performed KEGG pathway enrichment using `enrichKEGG()`, and applied `setReadable()` to convert enriched gene IDs back to symbols for better interpretability:
```{r kegg-enrich, collapse=FALSE, message=FALSE, warning=FALSE}
options(timeout = 300)
kegg_enrich <- enrichKEGG(gene     = gene_id$ENTREZID,
                          organism = "hsa",
                          keyType  = "kegg")

kegg_enrich <- setReadable(kegg_enrich, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

head(kegg_enrich)
```

The `enrichplot` package is fully compatible with `clusterProfiler` and provides a rich set of visualization methods. Below we demonstrated several common plot types (dot plot, bar plot, and concept network plot):
```{r kegg-visualize, collapse=FALSE, fig.width=10, fig.height=18, fig.align='center', out.width='95%'}
library(enrichplot)
library(aplot)

p1 <- dotplot(kegg_enrich)
p2 <- barplot(kegg_enrich)
p3 <- cnetplot(kegg_enrich)

plot_list(p1, p2, p3, ncol = 1)
```

## GO enrichment analysis

[Gene Ontology (GO)](http://geneontology.org/) is a widely used framework that provides a structured, controlled vocabulary for annotating gene functions. It consists of three orthogonal ontologies:

- **BP** (Biological Process): biological objectives to which a gene contributes (e.g., apoptosis, signal transduction).
- **CC** (Cellular Component): cellular locations where a gene product is active (e.g., nucleus, membrane).
- **MF** (Molecular Function): molecular activities of a gene product (e.g., kinase activity, DNA binding).

GO enrichment analysis tests whether GO terms are over-represented in a gene set relative to the genome background, thereby revealing the biological functions and cellular contexts associated with the input genes.

Here, we reused the gene set and Entrez IDs prepared in the KEGG section above. Unlike `enrichKEGG()`, `enrichGO()` also supports `keyType = "SYMBOL"` directly via the `OrgDb` annotation, but for consistency we continued to use Entrez IDs.

```{r go-enrich, collapse=FALSE, message=FALSE, warning=FALSE}
go_bp <- enrichGO(gene     = gene_id$ENTREZID,
                  OrgDb    = org.Hs.eg.db,
                  ont      = "BP",
                  keyType  = "ENTREZID",
                  readable = TRUE)
head(go_bp)
```

To perform enrichment for **Cellular Component (CC)**, **Molecular Function (MF)**, or **all three categories simultaneously**, simply change the `ont` parameter:

```{r go-cc-mf, collapse=FALSE, message=FALSE, warning=FALSE}
## Cellular Component (CC)
go_cc <- enrichGO(gene = gene_id$ENTREZID, OrgDb = org.Hs.eg.db,
                  ont = "CC", keyType = "ENTREZID", readable = TRUE)

## Molecular Function (MF)
go_mf <- enrichGO(gene = gene_id$ENTREZID, OrgDb = org.Hs.eg.db,
                  ont = "MF", keyType = "ENTREZID", readable = TRUE)

## All GO categories (ALL)
go_all <- enrichGO(gene = gene_id$ENTREZID, OrgDb = org.Hs.eg.db,
                   ont = "ALL", keyType = "ENTREZID", readable = TRUE)
```

Similarly, GO enrichment results can be visualized with `enrichplot` (for further details, refer to the [enrichplot documentation](https://yulab-smu.top/biomedical-knowledge-mining-book/)). For the combined `ont = "ALL"` result, a faceted bar plot provides a convenient comparison across BP, CC, and MF:

```{r go-visualize, fig.width=10, fig.height=10, fig.align='center', out.width='95%', collapse=FALSE}
p_all <- barplot(
  go_all,
  x = "Count",
  showCategory = 10,
  split = "ONTOLOGY"
) +
  aes(fill = ONTOLOGY) +
  scale_fill_brewer(palette = "Set2") +
  enrichplot::autofacet(by = "row", scales = "free")

p_all + scale_y_discrete()
```

## Advanced visualization

`TCMDATA` provides several publication-ready visualization functions tailored for enrichment results in network pharmacology research.

### Lollipop plot

The lollipop plot offers an alternative to the standard dot/bar plot, combining line segments with terminal points to display enrichment significance and gene counts. `gglollipop()` is built on top of `enrichplot` and shares compatible grammar:

```{r lollipop, collapse=FALSE, fig.width=10, fig.height=8, fig.align='center', out.width='95%'}
p_lollipop <- gglollipop(go_bp, line.col = "skyblue", 
                         line.type = "dashed", 
                         palette = "Spectral", top_n = 10) + 
  theme(strip.text.y = element_text(angle = 0)) 

p_lollipop
```

For the combined `ont = "ALL"` result, lollipop plots can be faceted by ontology to compare enriched terms across BP, CC, and MF:

```{r go-lollipop, collapse=FALSE, fig.width=10, fig.height=10, fig.align='center', out.width='95%'}
p_lollipop_all <- gglollipop(go_all, line.col = "orange", 
                             split = "ONTOLOGY",line.type = "dashed", 
                             palette = "PiYG", top_n = 5, show_count = FALSE) + 
  facet_grid(ONTOLOGY ~ ., scales = "free_y") +
  theme(strip.text.y = element_text(angle = 0)) 

p_lollipop_all
```

### Dot-sankey plot

The dot-sankey plot combines a **Sankey diagram** (showing gene–pathway alluvial flows) with a **dot plot** (encoding significance and gene count) in a single panel. This dual-view layout facilitates simultaneous inspection of which genes contribute to each pathway and how significant each enrichment is:

```{r dot-sankey, collapse=FALSE, fig.width=13, fig.height=15, fig.align='center', out.width='100%', warning=FALSE, message=FALSE}
ggdot_sankey(kegg_enrich, n = 8, dot_x_var = "GeneRatio")
```

### GO bar plot

`go_barplot()` provides a colored bar chart specifically designed for `enrichGO(ont = "ALL")` results, where bars are grouped and colored by ontology category (BP, CC, MF). These figures are commonly used in network pharmacology publications for presenting GO enrichment results:

```{r go-bar, collapse=FALSE, fig.width=10, fig.height=8, fig.align='center', out.width='95%', warning=FALSE, message=FALSE}
go_barplot(go_all)
```

### GO circle plot

`gocircle_plot()` renders a multi-ring circular plot using `circlize` and `ComplexHeatmap`, encoding four layers of information in a compact layout: ontology category, term size with $-\log_{10}(p_{\text{adjust}})$, up/down-regulated gene counts, and RichFactor:

```{r go-circle, collapse=FALSE, fig.width=10, fig.height=10, fig.align='center', out.width='80%', warning=FALSE, message=FALSE}
gocircle_plot(go_all, top = 5)
```

## Gene Set Enrichment Analysis (GSEA)

Unlike ORA, which requires a pre-defined list of significant genes, Gene Set Enrichment Analysis (GSEA) uses the **entire ranked gene list** to detect coordinated changes in gene sets. This approach avoids arbitrary significance cutoffs and is more sensitive to subtle but consistent expression shifts across a pathway.

The typical GSEA workflow requires a **named numeric vector** (e.g., log2 fold change or Wald statistic), sorted in decreasing order, with gene identifiers as names. Here, we use the built-in `deg_earlydn` dataset from `TCMDATA` to construct such a ranked list.

### Prepare ranked gene list

```{r gsea-prepare, collapse=FALSE, message=FALSE, warning=FALSE}
data(deg_earlydn)

## convert gene symbols to Entrez IDs
deg_ids <- bitr(deg_earlydn$names,
                fromType  = "SYMBOL",
                toType    = "ENTREZID",
                OrgDb     = org.Hs.eg.db)

## merge with DESeq2 results and build ranked list
deg_merged <- merge(deg_earlydn, deg_ids, by.x = "names", by.y = "SYMBOL")
gene_rank  <- sort(setNames(deg_merged$log2FoldChange, deg_merged$ENTREZID),
                   decreasing = TRUE)

head(gene_rank)
```

### KEGG GSEA

```{r gsea-kegg, collapse=FALSE, message=FALSE, warning=FALSE}
options(timeout = 300)
gsea_kegg <- gseKEGG(geneList = gene_rank,
                     organism = "hsa",
                     eps      = 0,
                     pvalueCutoff = 0.05)

gsea_kegg <- setReadable(gsea_kegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
head(gsea_kegg)
```

The classic GSEA running-score plot visualizes the enrichment profile of selected pathways. `gseaplot2()` from `enrichplot` supports displaying multiple gene sets simultaneously:

```{r gsea-kegg-plot, fig.width=10, fig.height=6, fig.align='center', out.width='95%', collapse=FALSE, message=FALSE}
gseaplot2(gsea_kegg, geneSetID = 1:3, pvalue_table = TRUE)
```

### GO GSEA

```{r gsea-go, collapse=FALSE, message=FALSE, warning=FALSE}
gsea_go <- gseGO(geneList = gene_rank,
                 OrgDb    = org.Hs.eg.db,
                 ont      = "BP",
                 eps      = 0,
                 pvalueCutoff = 0.05)

gsea_go <- setReadable(gsea_go, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
head(gsea_go)
```

A ridge plot provides a density overview of enrichment score distributions across top GO terms:

```{r gsea-go-plot, fig.width=10, fig.height=8, fig.align='center', out.width='95%', collapse=FALSE, message=FALSE}
ridgeplot(gsea_go, showCategory = 15)
```