# Machine learning–based key target screening {#ml-analysis}

In network pharmacology, candidate targets derived from compound–target mapping and PPI network analysis often number in the dozens to hundreds. Identifying a concise set of **key targets**, which are the most predictive of the disease phenotype, is essential for downstream experimental validation and mechanistic interpretation. **Machine learning (ML)** offers a data-driven, unbiased approach to this task: by training classifiers on transcriptomic profiles and evaluating feature importance, ML methods can distill large candidate pools into biologically meaningful gene signatures.

A growing number of publications employ a **multi-algorithm consensus** strategy, in which several complementary ML methods (e.g., LASSO, Random Forest, SVM-RFE, XGBoost) are applied in parallel; genes consistently selected across methods are reported as the final key targets. This consensus approach mitigates algorithm-specific biases and enhances the robustness of the selected biomarkers.

`TCMDATA` provides an **ML module** that implements this workflow end-to-end: from data preparation and model fitting through consensus extraction and publication-ready visualization. Three evaluation modes are supported:

| Mode | Description | Use case |
|------|-------------|----------|
| **A** | All samples used for training; performance assessed via out-of-fold (OOF) or out-of-bag (OOB) predictions | Single dataset; most common in network pharmacology studies |
| **B** | Internal random train/test split | When an independent cohort is unavailable but held-out validation is desired |
| **C** | External validation on an independent test set | When a separate cohort (e.g., different GEO series) is available |

---

## Introduction
### Example dataset {#ml-data}

Throughout this chapter we use the built-in `covid19` dataset, which contains RNA-seq expression profiles from 100 COVID-19 patients (GSE157103). Samples are labelled as ICU (intensive care unit) or NonICU, providing a clinically relevant binary classification task.

```{r ml-load-data, eval=TRUE, collapse=FALSE}
library(TCMDATA)

data("covid19")
str(covid19, max.level = 1)
```

We extract the expression matrix (genes × samples) and the group labels. To simulate a realistic candidate gene list, we select the **top 50 most variable genes** by inter-sample variance — in practice these would be the candidate targets obtained from upstream PPI / compound–target analysis:

```{r ml-prepare-data, eval=TRUE, collapse=FALSE}
expr_mat <- as.matrix(covid19$expr)
group <- covid19$group_info$group

## Select top 50 most variable genes by row variance
gene_var <- apply(expr_mat, 1, var)
genes <- names(sort(gene_var, decreasing = TRUE))[1:50]

cat("Expression matrix:", nrow(expr_mat), "genes ×", ncol(expr_mat), "samples\n")
cat("Group levels:", levels(group), "\n")
cat("Top 50 high-variance genes selected:", length(genes), "\n")
cat(genes, sep = ", ")
```

---

### Data preparation {#ml-prepare}

`prepare_ml_data()` converts a gene × sample expression matrix into a structured `tcm_ml_data` object that all downstream modelling functions accept. It handles level reordering (so that the positive class is always `levels[1]`), zero-variance gene removal, and syntactic name cleaning.

For Mode A (full cross-validation, no hold-out), simply pass the matrix, group labels, and the desired positive class:

```{r ml-mode-a-data, eval=TRUE, collapse=FALSE}
ml_data <- prepare_ml_data(
  expr_mat,
  group,
  positive_class = "ICU",
  genes = genes)
```

| Parameter | Description | Default |
|-----------|-------------|---------|
| `expr_mat` | Numeric matrix (genes × samples) | — |
| `group` | Factor or character vector of class labels (length = `ncol(expr_mat)`) | — |
| `positive_class` | Which level to treat as the positive class | `NULL` (alphabetical) |
| `genes` | Character vector of candidate genes to retain | `NULL` (all genes) |
| `split` | Logical; `TRUE` activates Mode B (internal split) | `FALSE` |
| `train_ratio` | Training fraction for Mode B | 0.7 |
| `test_expr` / `test_group` | External test data for Mode C | `NULL` |

---

## Mode A: full cross-validation {#ml-mode-a}

In Mode A, **all samples** participate in model training. Model performance is assessed using out-of-fold (OOF) predictions from cross-validation (LASSO, SVM-RFE, XGBoost) or out-of-bag (OOB) estimates (Random Forest). This is the most widely used setup in network pharmacology ML screening, as it maximises the training data and avoids random split bias.

### LASSO regression {#ml-lasso}

The Least Absolute Shrinkage and Selection Operator (LASSO)<sup>[2]</sup> applies an $L_1$ penalty to logistic regression, driving irrelevant coefficients exactly to zero and thereby performing automatic feature selection. The penalty strength $\lambda$ is selected via cross-validation; `ml_lasso()` supports both `lambda.min` (lowest CV error) and `lambda.1se` (most parsimonious model within one standard error), with the latter being the default.

$$\min_{\beta} \left\{ -\frac{1}{n} \sum_{i=1}^{n} \ell(y_i, \beta_0 + x_i^T \beta) + \lambda \|\beta\|_1 \right\}$$

```{r ml-lasso-fit, eval=TRUE, collapse=FALSE}
res_lasso <- ml_lasso(ml_data, cv_folds = 10, seed = 2025)
res_lasso
```

#### CV error curve

The cross-validation curve shows the AUC (or deviance) as a function of $\log(\lambda)$. The two vertical dashed lines mark `lambda.min` and `lambda.1se`:

```{r ml-lasso-cv, eval=TRUE, fig.width=7, fig.height=5, fig.align='center', out.width='80%', collapse=FALSE}
plot_enet_cv(res_lasso)
```

#### Coefficient path

The coefficient path traces how each gene's coefficient evolves along the regularisation path. Setting `top_n = 20` restricts the display to the 20 features with the largest absolute coefficients:

```{r ml-lasso-path, eval=TRUE, fig.width=7, fig.height=5, fig.align='center', out.width='80%', collapse=FALSE}
plot_enet_path(res_lasso, top_n = 20)
```

#### Coefficient bar chart

A horizontal bar chart coloured by sign provides an intuitive summary of the selected features and their directional effects:

```{r ml-lasso-coefs, eval=TRUE, fig.width=7, fig.height=5, fig.align='center', out.width='80%', collapse=FALSE}
plot_enet_coefs(res_lasso)
```

---

### Random Forest and Boruta {#ml-rf}

Random Forest (RF)<sup>[3]</sup> constructs an ensemble of decision trees, each trained on a bootstrap sample of the data. Feature importance is quantified by the mean decrease in Gini impurity (or accuracy) across all trees. To move from importance ranking to formal feature *selection*, `ml_rf()` employs the **Boruta** algorithm<sup>[4]</sup>, which compares each real feature's importance against that of randomly shuffled "shadow" copies; only features that consistently outperform their shadows are confirmed as relevant.

The pipeline proceeds in three phases:

1. Train an RF on **all** candidate genes to obtain importance scores.
2. Run Boruta all-relevant feature selection (tentative features resolved via `TentativeRoughFix`).
3. *(Optional, default)* Re-fit an RF on the **selected** genes only, so that OOB metrics honestly reflect the chosen subset.

```{r ml-rf-fit, eval=TRUE, collapse=FALSE}
res_rf <- ml_rf(ml_data, n_trees = 500, seed = 2025)
res_rf
```

#### Boruta importance plot

The Boruta plot shows the median importance of each feature, coloured by selection decision (Confirmed / Tentative / Rejected). The dashed vertical line marks the shadow maximum threshold:

```{r ml-rf-boruta, eval=TRUE, fig.width=7, fig.height=6, fig.align='center', out.width='80%', collapse=FALSE}
plot_rf_boruta(res_rf, top_n = 25)
```

#### Variable importance (Gini)

A standard bar chart of Mean Decrease Gini for the top features in the refitted model:

```{r ml-rf-importance, eval=TRUE, fig.width=7, fig.height=5, fig.align='center', out.width='80%', collapse=FALSE}
plot_rf_importance(res_rf, top_n = 20)
```

---

### SVM-RFE {#ml-svm-rfe}

Support Vector Machine–Recursive Feature Elimination (SVM-RFE) iteratively trains a linear SVM and removes the feature(s) with the smallest weight magnitude $|w_j|^2$ at each step, following the backward elimination framework of Guyon et al.<sup>[1]</sup>. The optimal feature-subset size is determined by the cross-validated ROC (or accuracy) profile across a grid of subset sizes. `ml_svm_rfe()` wraps `caret::rfe()` with sensible defaults tailored for gene expression data.

```{r ml-svm-fit, eval=TRUE, collapse=FALSE, warning=FALSE}
res_svm <- ml_svm_rfe(ml_data, cv_folds = 5, cv_repeats = 3, seed = 2025)
res_svm
```

#### RFE accuracy profile

The profile plot visualises the CV metric (ROC or Accuracy) as a function of the number of features retained. The dashed red line indicates the selected optimal subset size:

```{r ml-svm-curve, eval=TRUE, fig.width=7, fig.height=5, fig.align='center', out.width='80%', collapse=FALSE}
plot_svm_rfe_curve(res_svm)
```

---

### XGBoost {#ml-xgboost}

eXtreme Gradient Boosting (XGBoost)<sup>[5]</sup> builds an ensemble of shallow decision trees sequentially, with each new tree correcting the residual errors of its predecessors. An internal $k$-fold CV with early stopping determines the optimal number of boosting rounds, preventing overfitting. Features are ranked by their cumulative information **Gain** across all trees.

```{r ml-xgb-fit, eval=TRUE, collapse=FALSE}
res_xgb <- ml_xgboost(ml_data, cv_folds = 5, seed = 2025)
res_xgb
```

#### Feature importance (Gain)

```{r ml-xgb-importance, eval=TRUE, fig.width=7, fig.height=5, fig.align='center', out.width='80%', collapse=FALSE}
plot_xgb_importance(res_xgb, top_n = 20)
```

---

### Consensus analysis {#ml-consensus}

The multi-algorithm consensus strategy identifies genes that are selected by at least `min_methods` models, thereby filtering out method-specific noise and retaining only robust biomarker candidates.

#### Running all methods together

`run_ml_screening()` provides a convenient one-call interface:

```{r ml-screening, eval=TRUE, collapse=FALSE, warning=FALSE}
ml_list <- run_ml_screening(
  ml_data,
  methods = c("lasso", "rf", "svm_rfe", "xgboost"),
  cv_folds = 5,
  seed = 2025
)
summary(ml_list)
```

#### Venn diagram

```{r ml-venn, eval=TRUE, fig.width=7, fig.height=6, fig.align='center', out.width='70%', collapse=FALSE, warning=FALSE}
plot_ml_venn(ml_list)
```

#### Extracting consensus genes

```{r ml-consensus-genes, eval=TRUE, collapse=FALSE}
consensus <- get_ml_consensus(ml_list, min_methods = 2)
cat("Consensus genes (≥ 2 methods):\n")
cat(consensus, sep = ", ")
```

#### ROC curves

The ROC curves visualise discrimination performance for each method. In Mode A, curves are derived from out-of-fold / out-of-bag predictions, providing an honest assessment without data leakage:

```{r ml-roc, eval=TRUE, fig.width=7, fig.height=6, fig.align='center', out.width='80%', collapse=FALSE}
plot_ml_roc(ml_list)
```

---

## Mode B: internal train/test split {#ml-mode-b}

When a single dataset is available but held-out validation is desired, Mode B randomly partitions samples into training and test sets. This avoids the optimistic bias of purely cross-validated metrics while still operating within a single cohort.

```{r ml-mode-b, eval=TRUE, collapse=FALSE, warning=FALSE}
ml_data_B <- prepare_ml_data(
  expr_mat, group,
  positive_class = "ICU",
  genes = genes,
  split = TRUE,
  train_ratio = 0.7,
  seed = 2025
)

ml_list_B <- run_ml_screening(
  ml_data_B,
  methods = c("lasso", "rf", "svm_rfe", "xgboost"),
  cv_folds = 5,
  seed = 2025
)
```

The ROC curves are now evaluated on the held-out **test set**, yielding an independent estimate of classifier performance:

```{r ml-mode-b-roc, eval=TRUE, fig.width=7, fig.height=6, fig.align='center', out.width='80%', collapse=FALSE}
plot_ml_roc(ml_list_B)
```

```{r ml-mode-b-venn, eval=TRUE, fig.width=7, fig.height=6, fig.align='center', out.width='70%', collapse=FALSE, warning=FALSE}
plot_ml_venn(ml_list_B)
```

---

## Mode C: external validation {#ml-mode-c}

Mode C mirrors the most rigorous study design: the model is trained on one cohort and validated on an entirely independent dataset. `prepare_ml_data()` automatically intersects the gene spaces of the two matrices to ensure compatibility.

```{r ml-mode-c, eval=FALSE, collapse=FALSE}
## Example (not run): two separate GEO datasets
ml_data_C <- prepare_ml_data(
  train_expr, train_group,
  test_expr  = external_expr,
  test_group = external_group,
  positive_class = "Disease",
  genes = candidate_genes
)

ml_list_C <- run_ml_screening(ml_data_C)
plot_ml_roc(ml_list_C)    # ROC on external test set
plot_ml_venn(ml_list_C)
```

Because external validation is dataset-dependent, we illustrate it here with a simulated split from the same cohort:

```{r ml-mode-c-sim, eval=TRUE, collapse=FALSE, warning=FALSE}
set.seed(2025)
idx <- caret::createDataPartition(group, p = 0.7, list = FALSE)[, 1]

ml_data_C <- prepare_ml_data(
  expr_mat[, idx], group[idx],
  test_expr  = expr_mat[, -idx],
  test_group = group[-idx],
  positive_class = "ICU",
  genes = genes
)

ml_list_C <- run_ml_screening(
  ml_data_C,
  methods = c("lasso", "rf", "svm_rfe", "xgboost"),
  cv_folds = 5,
  seed = 2025
)
```

```{r ml-mode-c-roc, eval=TRUE, fig.width=7, fig.height=6, fig.align='center', out.width='80%', collapse=FALSE}
plot_ml_roc(ml_list_C)
```

---

## Post-hoc feature trimming {#ml-select}

After initial model fitting, `select_features()` allows fine-grained control over the number of retained genes without re-training the full model. This is particularly useful for Ridge regression and XGBoost, which do not inherently perform variable selection:

```{r ml-select, eval=TRUE, collapse=FALSE}
res_xgb_top10 <- select_features(res_xgb, top_n = 10)
cat("Original features:", length(res_xgb$genes), "\n")
cat("After trimming:  ", length(res_xgb_top10$genes), "\n")
cat("Top 10 genes:", paste(res_xgb_top10$genes, collapse = ", "), "\n")
```

---

## Assembling models manually {#ml-manual-list}

In some workflows you may wish to run individual model functions in separate
sessions—or selectively combine only a subset of methods—rather than calling
`run_ml_screening()` in one shot.  `create_tcm_ml_list()` assembles any
collection of `tcm_ml` objects into a `tcm_ml_list` that is fully compatible
with all downstream utilities (`plot_ml_roc()`, `plot_ml_venn()`,
`get_ml_consensus()`).

```{r ml-manual-prepare, eval=FALSE}
## prepare the same ml_data for every model
ml_data <- prepare_ml_data(expr_mat, group_vec)

## run whichever models you need
res_lasso <- ml_lasso(ml_data)
res_rf <- ml_rf(ml_data)
res_svm <- ml_svm_rfe(ml_data)
```

Pass the individual results as named arguments to `create_tcm_ml_list()`.
The names you provide become the method labels used in all downstream plots.
If you omit names, the function falls back to the `$method` slot stored inside
each object.

```{r ml-manual-create, eval=FALSE}
ml_list_custom <- create_tcm_ml_list(
  lasso = res_lasso,
  rf = res_rf,
  svm_rfe = res_svm)

## downstream analyses are identical to run_ml_screening() output 
# plot_ml_roc(ml_list_custom)                      
# plot_ml_venn(ml_list_custom)             
# get_ml_consensus(ml_list_custom, min_methods = 2) 
```

---

## Single-gene diagnostic analysis {#ml-gene-diag}

After identifying consensus genes via `get_ml_consensus()`, it is natural to
ask: *how well does each gene alone discriminate between the two groups?*
TCMDATA provides three companion functions for this purpose.

### AUC summary table

`get_gene_auc()` treats each gene's expression as a univariate predictor and
returns a tidy table with AUC, 95% DeLong CI, optimal cutoff (Youden's J),
sensitivity, specificity and a test against AUC = 0.5.

```{r ml-gene-auc, eval=TRUE}
consensus <- get_ml_consensus(ml_list, min_methods = 2)
auc_tbl   <- get_gene_auc(consensus, ml_data = ml_data)
head(auc_tbl)
```

Both `ml_data` (from `prepare_ml_data()`) and raw `expr_mat + group` are
accepted.  When `ml_data` is Mode B / C, the **test set** is used
automatically; for Mode A the training set is used.

### Single-gene ROC curves

`plot_gene_roc()` overlays one ROC curve per gene on a single panel:

```{r ml-gene-roc, eval=TRUE, fig.width=6, fig.height=5}
plot_gene_roc(consensus, ml_data = ml_data)

## Optional: show 95% CI bands
plot_gene_roc(consensus, ml_data = ml_data, show_ci = TRUE)
```

The legend lists genes in descending AUC order, and the full AUC table is
available via `attr(p, "auc_table")`.

### Two-group expression boxplot

`plot_gene_boxplot()` draws a faceted boxplot with statistical test annotation
(Wilcoxon rank-sum by default), jittered data points and optional violin
overlay — publication-ready with zero extra dependencies.

```{r ml-gene-box, eval=TRUE, fig.width=8, fig.height=6}
## Default: Wilcoxon test, p-value as number
plot_gene_boxplot(consensus, ml_data = ml_data)

## Stars instead of numeric p, add violin overlay
plot_gene_boxplot(consensus, ml_data = ml_data,
                  violin = TRUE, p_label = "p.signif")

## Also works with raw expression matrix
# plot_gene_boxplot(consensus, expr_mat = expr_mat, group = group)
```

The underlying test statistics are stored in `attr(p, "test_table")`.

---

## Summary {#ml-summary}

The table below summarises the ML methods available in `TCMDATA` and their key characteristics:

| Method | Function | Selection mechanism | Key hyperparameters |
|--------|----------|--------------------|--------------------|
| LASSO | `ml_lasso()` | $L_1$ penalty drives coefficients to zero | `lambda_rule`, `cv_folds` |
| Elastic Net | `ml_enet()` | $L_1 + L_2$ penalty (configurable `alpha`) | `alpha`, `lambda_rule` |
| Ridge | `ml_ridge()` | $L_2$ penalty; requires manual `top_n` cutoff | `top_n`, `lambda_rule` |
| Random Forest | `ml_rf()` | Boruta all-relevant selection on Gini importance | `n_trees`, `max_runs` |
| SVM-RFE | `ml_svm_rfe()` | Backward elimination by SVM weight $\|w_j\|^2$ | `kernel`, `cv_folds`, `cv_repeats` |
| XGBoost | `ml_xgboost()` | Gain-based ranking; requires manual `top_n` cutoff | `nrounds`, `max_depth`, `eta` |

---

## References

1. Guyon, I., Weston, J., Barnhill, S., & Vapnik, V. (2002). Gene selection for cancer classification using support vector machines. *Machine Learning*, 46, 389–422.
2. Tibshirani, R. (1996). Regression shrinkage and selection via the lasso. *Journal of the Royal Statistical Society: Series B*, 58(1), 267–288.
3. Breiman, L. (2001). Random forests. *Machine Learning*, 45, 5–32.
4. Kursa, M. B., & Rudnicki, W. R. (2010). Feature selection with the Boruta package. *Journal of Statistical Software*, 36(11), 1–13.
5. Chen, T., & Guestrin, C. (2016). XGBoost: A scalable tree boosting system. In *Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining* (pp. 785–794).

---

## Session Information {#ml-session-info}

```{r session-info, collapse=TRUE}
sessionInfo()
```
