# AI Module {#ai-module}

`TCMDATA` includes an AI module built on top of [aisdk](https://github.com/YuLab-SMU/aisdk), providing three layers of AI capability: **interpretation** (one-shot analysis of R objects), **agent** (multi-turn interactive analysis with tool use), and **skill** (domain-specific workflow knowledge). This chapter demonstrates the key features of each layer.

```{r ai-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

## Prerequisites

The AI module requires both `TCMDATA` and `aisdk`:

```{r install, eval=FALSE}
devtools::install_github("YuLab-SMU/aisdk")
library(aisdk)
library(TCMDATA)
```

Model configuration is handled through `tcm_setup()` and only needs to be called once per session. User must set their own API key and provider to use AI module in this package.

```{r config, eval=FALSE}
tcm_setup(
  provider = "openai",
  api_key  = "sk-xxx",
  model    = "gpt-4o", 
  base_url = "https://xxx/v1", # if necessary 
  save     = TRUE,
  test     = TRUE)
```

## Interpretation layer

The interpretation layer provides one-shot AI analysis of R objects and free-text queries, with no tool use or multi-turn conversation. The primary interface is `tcm_interpret()`.

### Free-text interpretation

```{r text-demo, eval=FALSE}
txt <- tcm_interpret(
  "Please introduce the major pharmacological functions of Huangqi (Astragalus membranaceus), focusing on immunoregulation.",
  language = "en")
cat(txt)
```


### Interpreting enrichment objects

`tcm_interpret()` can directly accept a `clusterProfiler` enrichment object. The package automatically compresses the relevant terms and genes into a compact representation before sending to the model:

```{r enrich-object, eval=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)

lz_targets <- search_herb("lingzhi", type = "Herb_pinyin_name")$target
lz_targets <- sample(unique(na.omit(lz_targets)), 100)

bp <- enrichGO(
  gene    = lz_targets,
  ont     = "BP",
  OrgDb   = org.Hs.eg.db,
  keyType = "SYMBOL")

tcm_interpret(
  bp,
  prompt   = "Summarise the main biological implications of this Lingzhi GO enrichment.",
  language = "en")
```


```{}
=== TCM AI Analysis: enrichment ===
  Model: openai:gpt-4o | Language: en | Audience: researcher

-- Summary --
The enrichment profile indicates that Lingzhi's putative targets cluster in innate
immune and inflammatory signaling, particularly TLR/IL-1/TNF-NF-κB pathways,
leukocyte adhesion, and matrix remodeling...

-- Key Findings --
  * Strong enrichment for cellular response to stress and external stimuli
  * Activation of innate immune and inflammatory pathways (IRAK1, NFKBIA, TNF, ADAM17)
  * Leukocyte adhesion and endothelial activation (VCAM1, MMP9)
  ...
```

### Interpreting PPI objects

PPI network objects with topological metrics computed by `compute_nodeinfo()` can also be directly interpreted:

```{r ppi-object, eval=FALSE}
data("demo_ppi")
ppi_graph <- compute_nodeinfo(demo_ppi)

tcm_interpret(
  ppi_graph,
  prompt   = "Characterise the network architecture and identify the most important hub nodes.",
  language = "en")
```

### Drafting result paragraphs

`draft_result_paragraph()` transforms an interpretation object into a publication-ready paragraph:

```{r draft-demo, eval=FALSE}
ai_res <- tcm_interpret(bp, language = "en")
draft  <- draft_result_paragraph(ai_res, language = "en")
cat(as.character(draft))
```


### Custom structured output

`tcm_interpret_schema()` allows user-defined output schemas for integration into downstream pipelines:

```{r custom-schema, eval=FALSE}
my_schema <- tcm_schema(
  summary     = tcm_field_string("A concise 2-3 sentence summary"),
  mechanism   = tcm_field_string("Core mechanistic interpretation"),
  key_targets = tcm_field_array("Most important targets"),
  confidence  = tcm_field_enum(c("high", "medium", "low"),
                               "Confidence level")
)

custom_res <- tcm_interpret_schema(
  bp,
  schema = my_schema,
  type   = "enrichment",
  prompt = "Focus on inflammation-related processes.")

print(custom_res)
```

Available field types: `tcm_field_string()`, `tcm_field_number()`, `tcm_field_boolean()`, `tcm_field_array()`, `tcm_field_enum()`.


## Agent layer

The agent layer adds **tool use** and **multi-turn conversation** on top of the interpretation engine. TCMDATA provides 30+ analysis tools (target search, enrichment, PPI, ML screening, compound lookup, visualization, etc.) that the agent can call autonomously.

### One-shot task: `tcm_agent()`

For a single analysis request, `tcm_agent()` automatically routes the query to the appropriate tools and returns the result:

```{r agent-oneshot, eval=FALSE}
# Simple herb target lookup — agent selects the right tool automatically
result <- tcm_agent("Search the targets of Huangqi (Astragalus)")
cat(result$text)

# The result also contains any generated artifacts
result$artifacts
```

The built-in **router** matches user queries to relevant tools using keyword patterns. For example, "enrichment" routes to GO/KEGG tools, "ppi" routes to network tools, and "machine learning" routes to ML screening tools. When multiple patterns match, tools are merged.

### Interactive session: `tcm_chat()`

`tcm_chat()` opens an interactive REPL for multi-turn exploratory analysis:

```{r agent-chat, eval=FALSE}
tcm_chat()
```

```
╔══════════════════════════════════════════════════╗
║  TCM-Pharmacist · Interactive Session            ║
║  Type /help for commands · /quit to exit         ║
╚══════════════════════════════════════════════════╝

[1] You > Search targets of Huangqi and sepsis, then compute intersection
 ✓ Route: target_lookup + disease_lookup (high)
 ✓ Tools called: search_herb_records → search_disease_targets → compute_target_intersection
 ✓ New artifacts: intersect_001

 Agent > Found 121 intersection targets between Huangqi and sepsis...

[2] You > Run GO and KEGG enrichment on the intersection
 ✓ Route: enrichment (high)
 ✓ Tools called: run_go_enrichment → run_kegg_enrichment
 ✓ New artifacts: enrich_001, enrich_002

 Agent > GO enrichment identified 245 significant BP terms...

[3] You > /artifacts
  intersect_001  gene_list    character[121]  2026-04-10 14:32:01
  enrich_001     enrichment   enrichResult    2026-04-10 14:32:15
  enrich_002     enrichment   enrichResult    2026-04-10 14:32:16

[4] You > /save 10x8
  ✓ Exported 3 artifacts to tcm_output/

[5] You > /quit
```

Key commands: `/help`, `/artifacts`, `/save [WxH]`, `/history`, `/stats`, `/quit`.

The session returns a list with `history` and `artifacts` for programmatic access:

```{r chat-capture, eval=FALSE}
res <- tcm_chat()
# After /quit:
res$artifacts  # all generated artifacts
res$history    # full conversation history
```

### Programmatic agent: `create_tcm_task_agent()`

For scripted workflows, create a reusable agent and execute tasks programmatically:

```{r agent-programmatic, eval=FALSE}
agent <- create_tcm_task_agent()

r1 <- run_tcm_task(agent, "Search Huangqi targets and sepsis targets, compute intersection")
r2 <- run_tcm_task(agent, "Run GO enrichment on the intersection genes")
r3 <- run_tcm_task(agent, "Build PPI network and rank hub genes")

# Each result contains: $text, $artifacts, $tool_calls
r3$text
```

### Artifact management

All analysis results are automatically stored as **artifacts** with generated IDs (e.g., `enrich_001`, `ppi_002`). The agent can reference artifacts in subsequent turns, and artifacts can be managed via dedicated functions:

```{r artifacts-demo, eval=FALSE}
list_tcm_artifacts()     # list all artifacts
load_tcm_artifact("enrich_001")  # retrieve the R object
clear_tcm_artifacts()    # clear all
```

## Skill layer

Skills are domain-knowledge packages that guide the agent through complex multi-step workflows. Unlike tools (which perform specific operations), skills provide **strategic context** — what to do, in what order, and what to watch out for.

### Built-in skills

TCMDATA ships with three skills:

| Skill | Purpose |
|-------|---------|
| `tcm-network-pharmacology` | Guides the full network pharmacology workflow (target retrieval → intersection → PPI → enrichment → validation → report). Only activates when user explicitly requests systematic analysis |
| `analysis-preferences` | Background constraint layer. Sets default parameters (e.g., p-value cutoffs, PPI score thresholds, visualization defaults) and quality standards. Active on every turn |
| `skill-creator` | Meta-skill for creating new custom skills |

### How skills work

When the agent detects a request that matches a skill's trigger condition, the skill's instructions are injected into the conversation context. For example, asking "Help me do a network pharmacology analysis of Huangqi treating sepsis" activates the `tcm-network-pharmacology` skill, which guides the agent through:

- **Phase 1**: Target collection (herb targets + disease targets → intersection)
- **Phase 2**: Network & enrichment (PPI construction, GO/KEGG analysis, hub gene ranking)
- **Phase 3**: Expression validation (WGCNA, ML screening, DEG integration — if data available)
- **Phase 4**: Single-cell validation (if scRNA-seq data available)
- **Phase 5**: Literature validation, cross-database verification, and report generation

Importantly, the skill is scope-aware: asking "Just search the targets of Huangqi" will only run the relevant step, not the full pipeline.

### Creating custom skills

Use the `skill-creator` skill to define new skills. A minimal skill only needs a `SKILL.md` file with YAML frontmatter:

```{r custom-skill, eval=FALSE}
# Example: ask the agent to create a new skill
tcm_chat()
# > Help me create a skill for molecular docking analysis
```

A skill directory can optionally contain `references/` (detailed documentation), `scripts/` (executable scripts), and `assets/` (templates).

### Managing skills

By default, the agent uses skills bundled with the package. To customize preferences or add new skills, initialize a local skills directory:

```{r skill-init, eval=FALSE}
# Copy all bundled skills to ./tcm_skills/ for customization
tcm_init_skills()
```

This creates a local `tcm_skills/` directory containing all three built-in skills. You can then:

```{r skill-edit, eval=FALSE}
# Edit analysis preferences (e.g., change default p-value cutoff)
file.edit("tcm_skills/analysis-preferences/SKILL.md")

# Add a new skill created by skill-creator
# Just place the skill directory under tcm_skills/:
#   tcm_skills/my-new-skill/SKILL.md

# Check which skills directory is currently active
tcm_skill_dir()

# Switch to a different skills directory
tcm_use_skills("path/to/other/skills")

# Reset to package defaults
tcm_reset_skills()
```

Once a local skills directory is active, all agent functions (`tcm_agent()`, `tcm_chat()`, `create_tcm_task_agent()`) will automatically use it.


## Session information

```{r ai-session-info}
sessionInfo()
```
